---
title: "R Day 3"
author: "Naomi Kaduwela"
date: "9/19/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Read a new dataset :: California Energy Data
```{r}
generation <- read.csv(here::here("data/generation.csv"),
                        stringsAsFactors = F)
imports <- read.csv(here::here("data/imports.csv"),
                        stringsAsFactors = F)
```
#Look at data
```{r}
str(generation)
str(imports)
```
#Fix date time formats
```{r}
library(lubridate)
generation$datetime <-as_datetime(generation$datetime)
class(generation$datetime)
head(generation$datetime)

imports$datetime <-as_datetime(imports$datetime)
class(imports$datetime)
head(imports$datetime)
```
#Wide vs long data :: need to make Generation data long by melting it on the date time variable so it abides by Tidy data standards
```{r}
head(generation)
library(reshape2)
long_gen <- melt(generation, id.vars = "datetime",
                 variable.name ="source",
                 value.name ="usage")
head(long_gen)
long_gen[order(long_gen$datetime)[1:20],]
```
#Merge Data on date time using original generation df and imports df
```{r}
merged_energy <- merge(generation, imports, by ="datetime")
dim(merged_energy)
head(merged_energy)
```
#Melt Merged Data
```{r}
long_merged_energy <- melt(merged_energy, id.vars = "datetime",
                 variable.name ="source",
                 value.name ="usage")
head(long_merged_energy)
```

#Playing with dplyr to select subsets of the merged data frame:
```{r}
library(dplyr)
tmp <- select(merged_energy, biogas, biomass, geothermal, solar)
names(tmp)

tmp <- select(merged_energy, contains("hydro"), starts_with("bio"))
names(tmp)

```
#Filter data with multiple conditions
```{r}
tmp <- filter(merged_energy, imports > 7000, natural_gas<7000)
nrow(tmp)
```
#Look at data again
```{r}
head(tmp)
```
#Mutate Data, to create new variables : Natural Log of usage, usage squared, usage to the third power
```{r}
tmp <- mutate(long_merged_energy, log_usage = log(usage), usage2 = usage^2, usage3 = usage^3)
head(tmp)
```
#Summarize data to a single value based on sum function : Total energy consumption
```{r}
summarize(long_merged_energy, total=sum(usage, na.rm=T))
```
#Summarize data to a single value based on mean function : mean energy consumption
```{r}
summarize(long_merged_energy, mean_cons=mean(usage, na.rm=T))
```
#Chaining functions with Pipe %>% "then" - creates an implicit data frame that takes next each operation. 
```{r}
long_merged_energy %>%
  filter(source == "geothermal") %>%
  select(-datetime) %>%
  mutate(log_usage = log(usage)) %>%
  summarize(mean_log_usage = mean(log_usage, na.rm=T))
```

#Summing with Pipe, can store to another variable 
```{r}
tmp <-
mereged_energy %>%
  select(-datetime) %>%
  mutate(total_usage = rowSums(.,na.rm=T)) %>%
  summarize(tota_usage = sum(tota_usage, na.rm=T))

```
#View merged_energy
```{r}
View(merged_energy)
```


#Piping: Select variables with hydro, create new var total_hydro which is the sum of all retained hydro vars 3. mean usage for total hydro 
```{r}
merged_energy %>%
  select(contains("hydro")) %>%
  mutate(total_hydro = rowSums(.,na.rm=T)) %>%
  summarize(mean_usage = mean(total_hydro, na.rm=T))
```
#Perform operations with Group by and summarize
```{r}
long_merged_energy %>%
  group_by(source) %>%
  summarize(sum_usage = sum(usage, na.rm=T))
```
#Perform operations with Group by and summarize
```{r}
gapminder <- read.csv(here::here("data/gapminder5.csv"))

gapminder %>%
  group_by(year) %>%
  summarize(mean_le = mean(lifeExp, na.rm=T),
            sd_lf = sd(lifeExp, na.rm=T))
```
#Mean by hour for small hydro, large hyrdo, biogas, and biomass - start with wide data set, melt, try to do with pipe

```{r}
merged_energy %>%
  select(datetime, contains("hyrdo"), contains("bio")) %>%
  melt(id.vars = "datetime", 
       variable.name = "source",
       value.name = "usage") %>%
  group_by(source) %>%
  summarize(mean_usage = mean(usage, na.rm=T))

```























































