---
title: "R Day 2"
author: "Naomi Kaduwela"
date: "9/18/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
here::here("data", "gapminder5.csv")
gapminder <- read.csv(here::here("data", "gapminder5.csv"))
str(gapminder)
```

## Converting Factors to Characters 

```{r}
gapminder$country <- as.character(gapminder$country)
gapminder$continent <- as.character(gapminder$continent)
str(gapminder)
```

## LOOPS: Create log variables with a FOR loop with GDP per capita and population called log_gdpPercap and log_pop
```{r}
obs <- 1:nrow(gapminder)
```
## Initialize for loop with for(i in vector)
```{r}
for(i in obs){
  gapminder[i,"log_gdpPercap"] <- log(gapminder[i,"gdpPercap"])
  gapminder[i, "log_pop"] <- log(gapminder[i,"pop"])
}
```
#Life expectancy increased over time? (mean by year)
```{r}
years <-unique(gapminder$year)

for(i in years){
  mean_le <- mean(gapminder$lifeExp[gapminder$year==i], na.rm=T)
  print(paste0(i, ": ", mean_le))
}
```

#Life expectancy increased over time? (mean by continent)
```{r}

continent <-unique(gapminder$continent)
for(i in continent){
  mean_le <- mean(gapminder$lifeExp[gapminder$continent==i], na.rm=T)
  print(paste0(i, ": ", mean_le))
}

```

## Nested FOR loops

```{r}
for (i in continent){
  print(paste0("Continent: ", i))
  for(j in years){
    std_le <- sd(gapminder$lifeExp[gapminder$continent == i & gapminder$year ==j], na.rm =T)
  print(paste0(j, ": ", std_le))
  }
}
```

# Apply - takes a matrix, instead of data frame 
```{r}
vars <- gapminder[, c("lifeExp", "pop", "gdpPercap")]
apply(vars, 2, mean)
```

## WHILE LOOP
```{r}
i <-1987
while (i<=2002){
  sd_le <- sd(gapminder$lifExp[gapminder$year ==i])
  print(paste0(i,": ", sd_le))
  i<- i+5
}
```

# IF
```{r}
set.seed(1)
random_year <- sample(years, 1)
if(random_year > 1997){
  print(random_year)
}else{
  print("sorry")
}
```
#FOR LOOP WITH IF ELSE - report mean population for years greater than or = 1987, print conditions not met as well
```{r}
threshold <- 1987
for(i in unique(gapminder$year)){
  tmp <- mean(gapminder$lifeExp[gapminder$year==i])
  
  if(tmp<threshold){
    print(paste("the year is < 1987 threshold"))
  } else{
    print(paste0("mean pop: ", tmp))
  }
}
```

## Writing Functions
```{r}
get_values <-
  function(df, variable = "continent"){
    vals <- unique(df[[variable]])
    print(paste0(variable, ": ", vals))
  }
```

## More coplex function

```{r}
report_mean_sd <-
  function(df, variable, country){
    var <- gapminder[[variable]][gapminder$country == country]
    m_le <- mean(var)
    sd_le <- sd(var)
    cat("Country", country, "\n Mean Life Expectancy: ", m_le, "\nSD Life Expectancy", sd_le)
  }

report_mean_sd(gapminder, "lifeExp", "Blugaria")
```
## create own function: reports mean, mediam, min, max for le for a continent in gapminder
```{r}
report_stats <-
  function(df, variable, continent){
    var <- gapminder[[variable]][gapminder$continent == continent]
    m_le <- mean(var)
    med_le <- median(var)
    min_le <- min(var)
    max_le <- max(var)
    cat("Continent", continent, 
        "\n Mean Life Expectancy: ", m_le, 
        "\n Median Life Expectancy", med_le,
        "\n Min Life Expectancy", min_le,
        "\n Max Life Expectancy", max_le
        )
  }

report_stats(gapminder, "lifeExp", "Oceania")

```

#Function with FOR loop
```{r}
viz_lm <-
  function(df, dv, iv, year){
    dat <- df[df[["year"]]==year,]
    y <- log(dat[[dv]])
    x <- log(dat[[iv]])
    fit <- lm(y-x)
    plot(y-x, main = year, 
         xlab = iv, ylab = dv)
    lines(x, predict(fit), col='blue')
  }

```


